\documentclass[review]{elsarticle}
\usepackage{lineno,hyperref,amssymb,graphicx,booktabs}

\modulolinenumbers[5]
\journal{Nuclear Instruments and Methods in Physics Research A}


% Note from Jeff: Our previous NMOR article was 20 pages in this
% format, which became about 6 pages in the final paper in NIMA.


%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
%\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}

\title{Sensitivity of Fields within Magnetically Shielded Volumes to
  Changes in Permeability}

\author[manitoba]{T. Andalib\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{andalibt@myumanitoba.ca}
\author[winnipeg,manitoba]{C.P. Bidinosti}
\author[winnipeg,manitoba]{R.R. Mammei}
\author[winnipeg,manitoba]{J.W. Martin}
\author[winnipeg]{A.~Bunch~of~other~People}

% Jeff to do: need to finalize author list.

\address[winnipeg]{Physics Department, The University of Winnipeg, 515 Portage Avenue, Winnipeg, MB, R3B 2E9, Canada}
\address[manitoba]{Department of Physics and Astronomy, University of Manitoba, Winnipeg, MB R3T 2N2, Canada}


\begin{abstract}
Future experiments seeking to measure the neutron electric dipole
moment (nEDM) require stable and homogeneous magnetic fields.  The
stability of the magnetic field within a magnetically shielded volume
is influenced by a number of factors.  In this paper, we study one of
these factors, which is the dependence of the internally generated
field on the permeability of the material.  We also provide
measurements of the temperature-dependence of the permeability of the
material, and indicate the extrapolation yet required to adequately
use these measurements to design future nEDM experiments.
\end{abstract}

\begin{keyword}
Magnetic Shielding \sep Neutron Electric Dipole Moment \sep Magnetic Field Stability
\end{keyword}

\end{frontmatter}

\linenumbers

\section{Introduction}

The next generation of neutron electric dipole moment (EDM)
experiments aim to measure the EDM $d_n$ with proposed precision
$\delta d_n\lesssim
10^{-27}$~e-cm~\cite{bib:nedm1,bib:nedm2,bib:nedm2.5,bib:nedm3,bib:nedm3.5,bib:nedm4,bib:nedm5,bib:nedm6,bib:nedm6.5}.
In the previous best experiment \cite{bib:pendlebury}, which discovered
$d_n<3\times 10^{-26}$~e-cm (90\% C.L), effects related to magnetic field
homogeneity and instability were found to dominate the systematic
error.  A detailed understanding of passive and active magnetic
shielding, magnetic field generation within shielded volumes, and
precision magnetometry is expected to be crucial to achieve the
systematic error goals for the next generation of experiments.  Much
of the R\&D effort for these experiments is focused on careful design
and testing of various magnetic shield geometries with precision
magnetometers~\cite{bib:brys,bib:afach,bib:fierlingerroom,bib:sturmthesis,bib:patton}.

%\begin{itemize}
%\item General requirements on field, stability of field, stability of gradient.
%\item List all possible factors that can affect field stability -
 % degaussing, vibrations, etc. possibly with additional references.
%\item State the problem addressed in this paper (changes in $\mu$) and
 % the result.
%\end{itemize}
The nEDM experiment at TRIUMF aims to determine $\delta d_n\sim
10^{-27}~e\cdot$cm. A DC magnetic field of typically $B_0=1~\mu$T is
used to provide the quantization axis for the UCN, normally directed
vertically in the apparatus.  The experimental cycle involves a Ramsey
sequence, where the UCN spins are oriented into the horizontal plane
by applying an oscillating horizontal $B_1$ field, followed by a free
precession period, then an additional application of a second $B_1$
pulse coherent in phase with the first.  The UCN are then drained from
the measurement cell and the number of UCN in both the spin-up state
and spin-down state are counted.  By varying the $B_1$ oscillation
frequency, the central resonant frequency may be deduce with high
precision.  Normally four fills of the measurement cell, with four
different $B_1$ frequencies are used to extract the resonant
frequency.  In order to extract $d_n$, the resonant frequency
measurement is conducted with an electric field $E$ oriented parallel
to the main $B_0$ field, and then again with $E$ oriented antiparallel
to the main $B_0$ field.  A problem in these experiments is if the
magnetic field drifts over the course of the measurement period, it
worsens the statistical precision with which $d_n$ can be determined.
For a $\delta d_n\sim 10^{-27}~e\cdot$cm measurement, the average
field over one fill of the measurement cell must be known to the 10~fT
level or better (which occurs typically on a timescale of 100~s), and
the field is seldom stable to this level.  For this reason,
experiments use a comagnetometer and/or surrounding atomic
magnetometers to correct the magnetic field to this
level~\cite{bib:nedm4,bib:afach,bib:brys}.  In order for
this correction technique succeed, gradient fields must be controlled
or corrected for at the $<1$~nT/m level.  Furthermore the stability of
the $B_0$ field must be controlled as well as possible, normally at
the 1-10~pT level over the UCN free-precession time.


The $B_0$ magnetic field generation system for a typical nEDM
experiment comprises a coil placed within a passively magnetically
shielded volume.  The passive magnetic shield is generally composed of
a multi-layer shield formed from thin shells of high-permeability
metal (mu-metal).  The outer layers of this shield are normally
cylindrical \cite{bib:nedm3.5, bib: nedm2
} or forming the walls of a magnetically
shielded room \cite{bib:altarev2014}.  The innermost magnetic shield is normally
a specially shaped shield, where the design of the coil in relation to
shield is carefully taken into account to achieve adequate
homogeneity.

%
The active compensation system will be used to reduce the external
%magnetic field at the measurement site to $\mu$T level and prevent the
%saturation of the passive magnetic shields.

mechanical and temperature changes of the passive magnetic
shielding~\cite{bib:voigt,bib:thiel}, and the demagnetization
procedure~\cite{bib:thiel,bib:fierlinger2016}, affect the stability of
the magnetic field within magnetically shielded
rooms.
%~\cite{bib:thiel,bib:altarev2014,bib:altarev2015,bib:voigt}.  
Active stabilization of the magnetic field surrounding magnetically
shielded rooms can also improve the internal
stability~\cite{bib:franke,bib:voigt, bib:afach}.  In nEDM
experiments, care must also be taken to stabilize the current supplied
to the $B_0$ coil~\cite{bib:brys}, and the coil must be stabilized
mechanically relative to the magnetic shielding as well.

% Jeff says: I question whether altarev2014 or 2015 actually say
% anything about magnetic stability.

% Taraneh answers: Altarev2014, they talk about their degaussing technique and
% that it leads to very low and homogeneous residual fields. They refer to 
% thiel's paper for its demonstration.
% Altarev 2015 is also about minimizing residual fields. Perhaps they don't
% belong to this section.

One additional affect, which is the subject of this paper, relates to
the fact that the $B_0$ coil in most nEDM experiments is magnetically
coupled to the innermost magnetic shield.  If the magnetic properties
of the innermost magnetic shield would change as a function of time,
it then results in a source of instability of $B_0$.  In the present
work, we sought to estimate this effect, and to characterize one
possible source of instability: changes the magnetic permeability
$\mu$ of the material with temperature.

While the senstivity of magnetic alloys to temperature variations has
been characterized in the past~\cite{bib:couderchon,bib:kruppvdm}, we
sought to make these measurements in regimes closer to the operating
parameters relevant to nEDM experiments.  For these alloys, it is also
known that the magnetic properties are set during the final annealing
process~\cite{bib:gupta,bib:bozorth,bib:kruppvdm}.  In this spirit we
performed our measurements on small witness cylinders that were
annealed using the process typically used in setting these properties.

% The temperature dependence of $\mu$ has been measured from two
% different approaches and the overall result found to be 0.1\%/K
% $<\frac{1}{\mu} \frac{d\mu}{dT}<$2.3\%/K.


\input{calculations.tex}




\section{Measurements of $\mu(T)$}

\subsection{Previous Measurements and their Relationship to nEDM Experiments\label{sec:previousmeasurement}}

Previous measurements of the temperature dependence of the magnetic
properties of high-permeability alloys have been summarized in
Refs.~\cite{bib:pfeifer,bib:bozorth,bib:couderchon}.  These
measurements are normally conducted using a sample of the material to
create a toroidal core, where a thin layer of the material is used in
order to avoid eddy-current and skin-depth
effects~\cite{bib:pfeifer,bib:kruppvdm}.  A value of $\mu$ is
determined by dividing the amplitude of the sensed $B$-field by the
amplitude of the driving AC $H$-field.  The value of $\mu$ is then
quoted either at or near its maximum attainable value by adjusting
$H$.  Depending on the details of the $B-H$ curve for the material in
question, this normally means that $\mu$ is quoted for $H$ being at or
near the coercivity of the
material~\cite{bib:couderchon,bib:kruppvdm}, often resulting in large
values up to $\mu_r=4\times 10^5$.

It is well known that $\mu$ measured for toroidal, thin metal wound
cores depends on the annealing process used, in particular on the
take-out or tempering temperature after the high-temperature portion
of the annealing process has been
completed~\cite{bib:pfeifer,bib:kruppvdm,bib:couderchon}.  Such
studies normally suggest a take-out temperature of 490-500$^\circ$C.
This ensures that the large $\mu_r=4\times 10^{5}$ is furthermore
maximal at room temperature.  Slight variations around room
temperature, and assuming the take-out temperature is not controlled
to better than a degree, imply a scale of possible temperature
variation of $\mu$ of approximately
$\frac{1}{\mu}\frac{d\mu}{dT}\simeq$0.3-1\%/K at room temperature would be
reasonable to assume~\cite{bib:couderchon,bib:kruppvdm}.

The chief challenge in applying these results to temperature stability
of nEDM experiments is that, when used as DC magnetic shielding, the
high-permeability alloys are usually operated for significantly
different parameters ($B$, $H$, and frequencies).

For example, when used in a shielding configuration, the effective
permeability is often measured to be typically $\mu_r=20,000$, in part
because the $H$-field in the material is below the DC coercivity
(typically $H_c=0.4$~A/m \cite{bib:kruppvdm}).  Another example is
that, as noted in Section~\ref{sec:theory_summary}, a more appropriate $H$ for
the innermost magnetic shield of an nEDM experiment used as a magnetic
flux return for the $B_0$ coil is 0.007~A/m, again well below the DC
coercivity.  Furthermore, the innermost magnetic shield of an nEDM
experiment is not operated in an AC mode, but rather a DC mode where
time (and temperature) stability of the DC $B_0$ field is the most
crucial parameter; i.e.~nEDM experiments are typically concerned with
slow drifts at $<0.01$~Hz timescales whereas the previously reported
$\frac{1}{\mu}\frac{d\mu}{dT}$ measurements are cycled at 50~Hz
frequencies.


The goal of our experiments was to develop techniques to characterize
the material properties of our own magnetic shields post-annealing, in
regimes more relevant to nEDM experiments.  We created a prototype
passive magnetic shield system in support of this (and other)
precision magnetic field research for the future nEDM experiment to be
conducted at TRIUMF.  The shield system is a four-layer mu-metal
shield formed from nested right-circular cylindrical shells with
endcaps.  The inner radius of the innermost shield is 18.44~cm which
is equal to its half length. The radii and half-lengths of the
progressively larger outer shields increase geometrically by a factor
of 1.27.  Each cylinder has two end-caps which possess a 7.5~cm
diameter central hole.  A stove-pipe of length 5.5~cm is place on each
hole was designed to minimize leakage of external fields into the
progressively shielded inner volumes.  The design is similar to
another smaller prototype shield discussed in
Ref.~\cite{bib:nmorpaper}.  The magnetic shielding factors of each of
the four cylindrical shells, and of various combinations of them, were
measured and found to be consistent with $\mu_r\sim 20,000-40,000$ for
3.5~$\mu$T applied external axial and transverse fields at frequencies
$<0.1$~Hz, where the uncertainty in $\mu_r$ comes a model uncertainty
resulting from incomplete inclusion of the geometry.

% Taraneh to do:  replace XXX with numbers

In our studies of the material properties of these magnetic shields,
two different approaches to measure temperature dependence of $\mu$
were pursued.  Both approaches involved experiments done using witness
cylinders, which are smaller open-ended cylinders (6'' in length) made
of the same material and annealed at the same time as the prototype
magnetic shields.  We therefore expect they have the same magnetic
properties as the larger prototype shields, and they have the
advantage of being smaller and easier to perform measurements with.



The two techniques used were:
\begin{enumerate}
\item measuring the axial shielding factor of the witness cylinder as
  a function of temperature, and
\item measuring the temperature-dependence of the slope of a minor B-H
  loop, using the witness cylinder as a transformer core, similar to
  previous measurements of the temperature dependence of $\mu$, but
  for parameters slightly closer to those encountered in nEDM
  experiments.
\end{enumerate}
We now discuss the details and results of each technique.

% Axial shielding factor measurements Section moved to axial.tex

\input{axial.tex}

% Transformer core measurements Section moved to transformer.tex

\input{transformer.tex}

\subsection{Summary of $\mu(T)$}

Each measurement (axial shielding factor measurement, or transformer
core measurement) is of the temperature dependence of an effective
$\mu$, done in an AC mode at a particular driving amplitude of $H(t)$.

In the axial shielding factor case, we found
0.6\%/K~$\lesssim\frac{1}{\mu}\frac{d\mu}{dT}\lesssim 2.7\%$/K, with
the measurement being conducted with a typical $H$-amplitude of 0.0001-0.001~A/m
and at a frequency of 1~Hz.  In the transformer core case, we found
0.1\%/K~$\lesssim\frac{1}{\mu}\frac{d\mu}{dT}\lesssim 2.1\%$/K, with
the measurement being conducted with a typical $H$-amplitude of
0.1~A/m and at a frequency of 1~Hz.

Our desire is to provide information relevant to EDM experiments where
the coil would couple to innermost magnetic shield in the experiment.
We think it would be reasonable to assume that for the parameters
relevant to EDM experiments ($H$ of 0.-0.007~A/m and frequencies
$<0.01$~Hz), the temperature dependence of the effective permeability
could be of similar scale to these measurements
i.e.~$\frac{1}{\mu}\frac{d\mu}{dT}$ ranging from 0.1 to 2.7\%/K.

\section{Summary and Conclusions}


Neutron EDM experiments are typically designed with the DC coil being
magnetically coupled to the innermost magnetic shield.  If the
magnetic permeability of the shield changes, this results in a change
in the field in the measurement region by an amount
$\frac{\mu}{B}\frac{dB}{d\mu}\sim 0.01$.  Temperature changes could
result in such changes, and it is those temperature changes we sought
to constrain by measurement.

The temperature dependence of $\mu$ has been measured by two different
techniques using open-ended mu-metal cylinders with 6 inches in
length.  We summarize the overall result as
0.1\%/K$<\frac{1}{\mu}\frac{d\mu}{dT}<$2.7\%/K, where the range is
driven in part by material properties of the different mu-metal
cylinders, and in part by fluctuations measured in the temperature
slopes in each measurement.

We provide this range with the following caveats:
\begin{itemize}
\item The measurement techniques rely on considerably larger
  frequencies and $H$-fields than those relevant to typical nEDM
  experiments.  For frequency, both techniques typically used a 1~Hz
  AC field, whereas for nEDM experiments the field is DC and stable at
  the 0.01~Hz level.  Furthermore, in one measurement technique the
  amplitude of $H$ was $\sim 0.001$~A/m and in the other was $\sim 0.1$~A/m,
  while for nEDM experiments $H\sim 0.-0.007$~A/m and is DC.
\item Both measurement techniques extract an effective $\mu$ that
  describes the slope of minor loops in B-H space.  A more correct
  treatment would include a more comprehensive accounting of
  hysteresis in the material, which is beyond the scope of this work.
\end{itemize}


Assuming our measurement of
0.1\%/K$<\frac{1}{\mu}\frac{d\mu}{dT}<$2.7\%/K and the generic EDM
experiment sensitivity of $\frac{\mu}{B}\frac{dB}{d\mu}\sim 0.01$
results in a temperature dependence of the magnetic field in a typical
EDM experiment of $\frac{dB}{dT}=10-270$~pT/K.  To achieve a goal
$\sim pT$ stability in the internal field for EDM experiments, the
temperature of the innermost magnetic shield in the EDM experiment
should then be controlled to 0.1-0.004~K level.  This could be a
serious problem for future nEDM experiments.  The dependence could be
reduced significantly by using self-shielded coils in the EDM
experiment, which would reduce the coupling to the innermost magnetic
shield.





%The next step is to measure $\mu(T)$ for the nEDM setup with different
%coil geometries to achieve the best stability.



%\begin{itemize}
%\item We determined $dB_0/d\mu$.  For reasonable parameters its $\sim$ XXX.  For self-shielded it's reduced by a factor of XXX for realistic windings.
%\item We measured $d\mu/dT$.  We constrained it to be in the range 0.1
 % to 2.3 percent per Kelvin.  Dominant errors are XXX for each method.
  %Neither method is for correct B, H, f.  Both methods are very
  %difficult to achieve consistent sub-percent determinations, but tend
  %to agree on sign and magnitude of problem.
%\item Results imply temperature stability to XXX Kelvin for stability
 % goal of XXX pT.  Show that this could be a dominant issue for
  %stability for future experiments.  Not addressed by many other
  %measurements which tend to focus on stability at zero field or
  %overall stability in full EDM experiment.
%\item Future work: build full EDM experiment.  Capability of various
 % internal coil geometries and best achievable temperature stability
  %are both important to success.
%\end{itemize}

%\section{To Do List}

%\begin{enumerate}
%\item Collect graphs and figures.
%\begin{enumerate}
%\item Analytical comparison reaction factor vs. $\mu$ for
 % sphere and/or cylinder with reasonable geometry/$\mu$.
%\item Same or additional comparison for more realistic geometry in
%  simulation.  May include self-shielded geometry.
%\item Axial shielding factor figure of setup.
%\item Axial shielding factor data.
%\item Transformer core Jiles-Atherton comparison?
%\end{enumerate}
%\item Write.
%\end{enumerate}

\section*{References}

\bibliography{mybibfile}
\begin{thebibliography}{00}

\bibitem{bib:nedm1} S. N. Balashov {\it et al.}, arXiv:0709.2428.

\bibitem{bib:nedm2} A. P. Serebrov {\it et al.}, JETP Lett. {\bf 99}, 4
  (2014).

\bibitem{bib:nedm2.5} A. P. Serebrov {\it et al.}, Physics Procedia {\bf
  17}, 251 (2011).

\bibitem{bib:nedm3} K. Kirch, AIP Conf. Proc., Vol. 1560, pp. 90-94
  (2013).

\bibitem{bib:nedm3.5} C. A. Baker, {\it et al.}, Physics Procedia {\bf
  17}, 159 (2011).

\bibitem{bib:nedm4} Y. Masuda, K. Asahi, K. Hatanaka, S.-C. Jeong,
  S. Kawasaki, R. Matsumiya, K. Matsuta, M. Mihara, and Y. Watanabe,
  Phys. Lett. A {\bf 376}, 1347 (2012).

\bibitem{bib:nedm5} I. Altarev, {\it et al.}, Nuovo Cim. C {\bf
  35}, 122 (2012).

\bibitem{bib:nedm6} R. Golub and S. K. Lamoreaux, Phys. Rept.  {\bf
  237}, 1 (1994).

\bibitem{bib:nedm6.5} T. M. Ito (the nEDM collaboration),
  J. Phys. Conf. Ser. {\bf 69} 012037, 2007.

\bibitem{bib:baker} C. A. Baker, {\it et al.}, Phys. Rev. Lett. {\bf
  97}, 131801 (2006).

\bibitem{bib:brys} T. Bry\'s, {\it et al.}, Nucl. Instrum. Meth. A
  {\bf 554}, 527 (2005).

\bibitem{bib:afach} S. Afach, {\it et al.}, J. Appl. Phys. 116, 084510 (2014).

\bibitem{bib:fierlingerroom} I. Altarev, {\it et al.}
  Rev. Sci. Instrum. {\bf 85}, 075106 (2014).

\bibitem{bib:sturmthesis} M. Sturm, Masterarbeit, T.U. Muenchen (2013).

\bibitem{bib:patton} B. Patton, E. Zhivun, D. C. Hovde, and D. Budker,
  Phys. Rev. Lett. 113, 013001 (2014).
\bibitem{bib:couderchon} G. Couderchon, J. F. Tiers, J. Magn. Magn. Mat. 26(1):196-214 (1982).

\bibitem{bib:bidinosti} C. P. Bidinosti, J. W. Martin, AIP Advances 4, 047135 (2014).

\bibitem{bib:smythe} W. R. Smythe, McGraw Hill (1950).

\bibitem{bib:ferraro} V. C. A. Ferraro, Athalone Press (1967).

\bibitem{bib:gupta} Kiran Gupta, K. K. Raina, S. K. Sinha, J. Alloys Compd, 429(1):357-364 (2007).

\bibitem{bib:knecht} A. Knecht, Dissertation, UZH Zurich (2009).

\bibitem{bib:kruppvdm} KruppVDM Magnifer 7904, MDS no. 9004 (2000).

\bibitem{bib:jiles} D. C. Jiles, D. L. Atherton,  J. Magn. Magn. Mat. 61(1-2):48-60, (1986).

\bibitem{bib:pendlebury} J. M. Pendlebury {\it et al.}, Phys. Rev. D 92, 092003 (2015).

\bibitem{bib:pfeifer} F. Pfeifer, C. Radeloff, J. Magn. Magn. Mat. 19 (1980) 190-207.

\bibitem{bib:thiel} F. Thiel {\it et al.}, Rev. Sci. Instrum. 78, 035106 (2007).

\bibitem{bib:altarev2014} I. Altarev {\it et al.},Rev. Sci. Instrum. 85, 075106 (2014).

\bibitem{bib:altarev2015} I. Altarev {\it et al.}, J. Appl. Phys. 117, 233903 (2015).

\bibitem{bib:voigt} J. Voigt {\it et al.}, Metrol. Meas. Syst. 20, 239 (2013).

\bibitem{bib:bozorth} R. M. Bozorth, Ferromagnetism, IEEE Press (1993).

\bibitem{bib:lockin} Stanford Reserach Systems, DSP lock-in amplifier, Model SR830 (2011)

\bibitem{bib:paperno-open-ended} E. Paperno, IEEE Trans. Magn. VOL. 35 NO. 5 (1999)

\bibitem{bib:fierlinger2016} Z. Sun {\it et al.}, J. App. Phys. 119, 193902 (2016)

\bibitem{bib:franke} B. Franke, Dissertation, ETH Zurich (2013)

\bibitem{bib:urankar} L. Urankar, R. Oppelt, IEEE Trans. Biomed. Eng. VOL. 43, NO. 7 (1996) 

\bibitem{bib:femm} D. Meeker, Finite Element Method Magnetics, FEMM, 4, 32 (2010)

\bibietm{bib:cpviolationwithoutstrangeness} I. B. Khriplovich, S. Lamoreaux, { \it CP violation without strangeness: electric dipole moments of particles, atoms, and molecules. }, Springer Science \& Business Media (2012)

\bibitem{bib:nmorpaper} J. W. Martin {\it et al.}, Nucl. Instrum. Meth. A 778:61-66 (2015) 

\bibitem{bib:bartman} Bartington Instruments, Operation Manual for Mag-03 Three-Axis Magnetic Field Sensors 
\end{thebibliography}

\end{document}
